---
title: "HW 2"
output: github_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl) # confirm this is the package he wants

```


## Problem 1

```{r}

# read in Mr. Trash Wheel data excluding the header and note column 
df_trash_data = read_excel("Data/HealthyHarborWaterWheelTotals2018-7-28.xlsx",
                           sheet = "Mr. Trash Wheel",
                           range = cell_cols("A:N"),
                           skip = 1)

# clean names
df_trash_data = df_trash_data %>% 
  janitor::clean_names()

# drop summary rows
df_trash_data = df_trash_data %>% 
  drop_na(dumpster)

# round sports balls field
df_trash_data = df_trash_data %>% 
  mutate(
    sports_balls = round(sports_balls, 0) %>% 
      as.integer()
  )

# read in 2017 precipitation data
df_precipitation_17 = read_excel("Data/HealthyHarborWaterWheelTotals2018-7-28.xlsx",
                                 sheet = "2017 Precipitation",
                                 range = "A2:B14")

# clean names
df_precipitation_17 = df_precipitation_17 %>% 
  janitor::clean_names()

# filter out missing values and add year feature
df_precipitation_17 = df_precipitation_17 %>% 
  drop_na(total) %>% 
  mutate(year = 2017)


# read in 2018 precipitation data
df_precipitation_18 = read_excel("Data/HealthyHarborWaterWheelTotals2018-7-28.xlsx",
                                 sheet = "2018 Precipitation",
                                 range = "A2:B14")

# clean names
df_precipitation_18 = df_precipitation_18 %>% 
  janitor::clean_names()

# filter out missing values and add year feature
df_precipitation_18 = df_precipitation_18 %>% 
  drop_na(total) %>% 
  mutate(year = 2018)


# join the datasets together
df_precipitation = rbind(df_precipitation_17, df_precipitation_18)

# change month values to character names
df_precipitation = df_precipitation %>% 
  mutate(
    month = month.name[month]
  )

```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in both resulting datasets, and give examples of key variables. For available data, what was the total precipitation in 2018? What was the median number of sports balls in a dumpster in 2017?

```{r}

df_precipitation %>% 
  group_by(year) %>% 
  filter(!(month %in% c("August", 
                        "September", 
                        "October", 
                        "November", 
                        "December"))) %>% 
  summarise(mean(total))

df_precipitation %>% 
  group_by(year) %>% 
  summarise(mean(total))

df_precipitation %>% 
  pivot_wider(names_from = year, values_from = total) %>% 
  mutate(difference = `2018` - `2017`)


```


## Problem 2

{First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.}


```{r}
# steps
# separate mon into year, month, and day
# combine prez fields

# load  and tidy pols data
df_pols_month = read_csv("Data/pols-month.csv") %>% 
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>% 
  mutate(month = month.name[month],
         president = case_when(
           prez_dem == 1 ~ "dem",
           prez_gop %in% c(1, 2) ~ "gop"
         )
  ) 
  
# check mapping of variables
#   note: 5 values of 2 for prez_gop; assumed these meant 1
df_pols_month %>% 
  count(president, prez_dem, prez_gop)

# drop prez and day fields
df_pols_month = df_pols_month %>% 
  select(-c(prez_dem, prez_gop, day)) %>% 
  arrange(year, month)


# load and clean snp data
df_snp = read_csv("Data/snp.csv") %>% 
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>% 
  mutate(month = month.name[month]) %>% 
  select(year, month, close) %>% 
  arrange(year, month)


# load and tidy unemployment data
df_unemployment = read_csv("Data/unemployment.csv") %>% 
  rename(year = Year) %>% 
  pivot_longer(-year, names_to = "month", values_to = "unemployment_rate") %>% 
  mutate(month = month.name[match(month, month.abb)]) %>% 
  arrange(year, month)

  
# join in s&p closing values
df_pols_month = left_join(df_pols_month, df_snp, 
                          by = c("year", "month")) 

# join in unemployment rates
df_pols_month = left_join(df_pols_month, df_unemployment, 
                          by = c("year", "month"))
  


```


{Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.}

{Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.}

{Join the datasets by merging snp into pols, and merging unemployment into the result.}

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).




## Problem 3

{Load and tidy the data. Note that, although these data may seem fairly well formatted initially, the names of a categorical predictor and the case structure of string variables changed over time; you’ll need to address this in your data cleaning. Also, some rows seem duplicated, and these will need to be removed (hint: google something like “dplyr remove duplicate rows” to get started).}

{Produce a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time; this should have rows for ethnicities and columns for year.}

{Produce a similar table showing the most popular name among male children over time.}

Finally, for male, white non-hispanic children born in 2016, produce a scatter plot showing the number of children with a name (y axis) against the rank in popularity of that name (x axis).


```{r}

# data checks
# df_baby_names %>% 
#   count(ethnicity)
# 
# temp <- df_baby_names %>% 
#   count(childs_first_name) %>% 
#   mutate(childs_first_name = tolower(childs_first_name)) %>% 
#   distinct()




# read and clean name data
df_baby_names = read_csv("Data/Popular_Baby_Names.csv") %>% 
  janitor::clean_names() %>% 
  arrange(year_of_birth) %>% 
  mutate(childs_first_name = tolower(childs_first_name),
         ethnicity = ethnicity %>% 
           recode("ASIAN AND PACI" = "ASIAN AND PACIFIC ISLANDER",
                  "BLACK NON HISP" = "BLACK NON HISPANIC",
                  "WHITE NON HISP" = "WHITE NON HISPANIC") %>% 
           tolower()) %>% 
  distinct()


# Produce a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time; this should have rows for ethnicities and columns for year
table_olivia = df_baby_names %>% 
  filter(childs_first_name == "olivia") %>% 
  select(-c(gender, count)) %>% 
  distinct() %>% 
  pivot_wider(names_from = year_of_birth, values_from = rank)

knitr::kable(table_olivia, 
             caption = "Popularity of Name Olivia for Females") 



# Produce a similar table showing the most popular name among male children over time.
table_pop_male_names = df_baby_names %>% 
  filter(gender == "MALE", rank == 1) %>% 
  select(-c(gender, count)) %>% 
  distinct() %>% 
  pivot_wider(names_from = year_of_birth, values_from = childs_first_name)

```


















