---
title: "HW 2"
output: github_document
---


```{r setup, include=FALSE}

library(tidyverse)
library(readxl)
library(viridis)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))

```



## Problem 1

```{r}

# read in Mr. Trash Wheel data excluding the header and note column
df_trash_data = read_excel("Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
                           sheet = "Mr. Trash Wheel",
                           range = cell_cols("A:N"),
                           skip = 1)

# clean names
df_trash_data = df_trash_data %>%
  janitor::clean_names()

# drop summary rows
df_trash_data = df_trash_data %>%
  drop_na(dumpster)

# round sports balls field and parse to numeric
df_trash_data = df_trash_data %>%
  mutate(
    sports_balls = round(sports_balls, 0) %>%
      as.integer()
  )

# read in 2017 precipitation data
df_precipitation_17 = read_excel("Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
                                 sheet = "2017 Precipitation",
                                 range = "A2:B14")

# clean names
df_precipitation_17 = df_precipitation_17 %>%
  janitor::clean_names()

# filter out missing values and add year feature
df_precipitation_17 = df_precipitation_17 %>%
  drop_na(total) %>%
  mutate(year = 2017)


# read in 2018 precipitation data
df_precipitation_18 = read_excel("Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
                                 sheet = "2018 Precipitation",
                                 range = "A2:B14")

# clean names
df_precipitation_18 = df_precipitation_18 %>%
  janitor::clean_names()

# filter out missing values and add year feature
df_precipitation_18 = df_precipitation_18 %>%
  drop_na(total) %>%
  mutate(year = 2018)


# join the datasets together
df_precipitation = rbind(df_precipitation_17, df_precipitation_18)

# change month values to character names
df_precipitation = df_precipitation %>%
  mutate(
    month = month.name[month]
  ) %>%
  select(year, month, total_precipitation = total)

```

The `Mr. Trash Wheel` data comprise `r ncol(df_trash_data)` columns and `r nrow(df_trash_data)` observations. Key features of this dataset include `weight_tons` and `volume_cubic_yards`, the weight and volume, respectively, of trash collected per month. The number of homes powered by incinerating the collected trash, `homes_powered`, is also an interesting feature that demonstrates the impact that Mr. Trash Wheel has on the Baltimore community. The dataset also contains quantities of certain itemized trash objects, such as the number of sports balls collected. In 2017 for example, the median number of balls collected per dumpster was `r df_trash_data %>% filter(year == 2017) %>% summarize(median = median(sports_balls))`.


The precipitation data describe the total precipitation in Baltimore per month. There are `r nrow(df_precipitation)` observations, representing precipitation for all months in 2017 and 2018. Notably, the total precipitation in 2018 was `r df_precipitation %>% filter(year == 2018) %>% summarize(sum = sum(total_precipitation))` inches, more than twice the `r df_precipitation %>% filter(year == 2017) %>% summarize(sum = sum(total_precipitation))` inches observed in 2017. In fact, 2017 precipitation only exceeded 2018 in 2 months.




## Problem 2

{First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.}


```{r}

# load  and tidy pols data
df_pols_month = read_csv("Data/pols-month.csv") %>%
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>%
  arrange(year, month) %>%
  mutate(month = month.name[month],
         president = case_when(
           prez_dem == 1 ~ "dem",
           prez_gop %in% c(1, 2) ~ "gop"
         )
  )

# check mapping of variables
#   note: 5 values of 2 for prez_gop; assumed these meant 1
df_pols_month %>%
  count(president, prez_dem, prez_gop)

# drop prez and day fields
df_pols_month = df_pols_month %>%
  select(-c(prez_dem, prez_gop, day))


# load and clean snp data
df_snp = read_csv("Data/snp.csv") %>%
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>%
  arrange(year, month) %>%
  mutate(month = month.name[month]) %>%
  select(year, month, close)


# load and tidy unemployment data
df_unemployment = read_csv("Data/unemployment.csv") %>%
  rename(year = Year) %>%
  pivot_longer(-year, names_to = "month", values_to = "unemployment_rate") %>%
  mutate(month = month.name[match(month, month.abb)])


# join in s&p closing values
df_pols_month_joined = left_join(df_pols_month, df_snp,
                                 by = c("year", "month"))

# join in unemployment rates
df_pols_month_joined = left_join(df_pols_month_joined, df_unemployment,
                                 by = c("year", "month"))

```




The `df_pols_month` dataset describes the political composition of the U.S. government at different points in time. There are values for the numbers of democrats and republicans in each of the houses of Congress and as governors, and indicators for the political affiliation of the president. There are `r nrow(df_pols_month)` per-month observations.

`df_snp` contains `r nrow(df_snp)` observations on the closing values of the Standard & Poor’s stock market index by year and month.

The `df_unemployment` dataset contains `r nrow(df_unemployment)` observations of the U.S unemployment rate by year and month.

When joined together, these datasets relate the political party composition of the U.S. government to the S&P index and unemployment rate, which are often used as recession indicators. The joined dataset comprises `r ncol(df_pols_month_joined)` features and `r nrow(df_pols_month_joined)` rows representing years `r df_pols_month_joined %>% summarize(min = min(year))` to `r df_pols_month_joined %>% summarize(max = max(year))`. The key values of `close` for the S&P index and `unemployment_rate` only have data starting in `r df_pols_month_joined %>% filter(!is.na(close)) %>% summarize(min = min(year))` and `r df_pols_month_joined %>% filter(!is.na(unemployment_rate)) %>% summarize(min = min(year))`, respectively.




## Problem 3



```{r}


# read and clean name data
df_baby_names = read_csv("Data/Popular_Baby_Names.csv") %>%
  janitor::clean_names() %>%
  arrange(year_of_birth) %>%
  mutate(childs_first_name = str_to_lower(childs_first_name),
         ethnicity = ethnicity %>%
           recode("ASIAN AND PACI" = "ASIAN AND PACIFIC ISLANDER",
                  "BLACK NON HISP" = "BLACK NON HISPANIC",
                  "WHITE NON HISP" = "WHITE NON HISPANIC") %>%
           tolower()) %>%
  distinct()

# check ethnicity
df_baby_names %>%
  count(ethnicity)

# check names: one name repeated from different capitalizations
df_baby_names %>%
  select(-c(count, rank)) %>%
  group_by(year_of_birth, gender, ethnicity, childs_first_name) %>%
  filter(n() > 1) %>%
  count(childs_first_name)

# Produce a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time; this should have rows for ethnicities and columns for year
table_olivia = df_baby_names %>%
  filter(childs_first_name == "olivia") %>%
  select(-c(gender, count)) %>%
  distinct() %>%
  pivot_wider(names_from = year_of_birth, values_from = rank)

knitr::kable(table_olivia,
             caption = "Popularity of name \"Olivia\" for females by ethnicity and year.")



# Produce a similar table showing the most popular name among male children over time.
# You need to create a data frame that has rows for ethnicities and columns for years, and each cell contains the most popular boy name for that ethnicity / year.
table_pop_male_names = df_baby_names %>%
  filter(gender == "MALE", rank == 1) %>%
  select(-c(gender, count)) %>%
  distinct() %>%
  pivot_wider(names_from = year_of_birth, values_from = childs_first_name)

knitr::kable(table_pop_male_names,
             caption = "Most popular male name by ethnicity and year.")



# For male, white non-hispanic children born in 2016, produce a scatter plot showing the number of children with a name (y axis) against the rank in popularity of that name (x axis)
ggp_names = df_baby_names %>%
  filter(gender == "MALE",
         ethnicity == "white non hispanic",
         year_of_birth == 2016)

ggplot(ggp_names, aes(x = rank, y = count)) +
  geom_point() +
  labs(title = "Count for Baby Names by Popularity",
       subtitle = "Male names in 2016 among white, non-hispanics",
       x = "Popularity rank",
       y = "Number of babies with given name")

```